Retrofitting Effect Handlers onto OCaml

### 2020-11-11 retro-concurrency.tex
Updates

Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\cite{WasmPropoal}. Project Loom~\cite{loom} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\ref{Farvardin}. In our experiments, we observed
that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
cache.

@@

We consider well-behaved Multicore OCaml programs to resume their continuations
\emph{exactly once}. This can be enforced by installing a finaliser on every
captured continuation that |discontinue|s the continuation during finalisation
and ignores the result:
\begin{lstlisting}
Gc.finalise (fun k ->
  try ignore (discontinue k Unwind) | _ -> ()) k
\end{lstlisting}
Multicore OCaml does not enforce this behaviour in order to minimise the cost
of using effect handlers.

@@

The interaction on non-local control flow and resources have been studied
extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
re-allocation) of resources when the non-local control leaves and enters back
into a context. |dynamic-wind| is considered to be complex and not quite what
one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge
is distinguishing returning exits from a computation closing over resources
from non-returning ones. Resource clean up needs to occur only during
non-returning exit. In Multicore OCaml, we assume that continuations are used
linearly, and hence, no clean up occurs when effects are performed but only
when an exception is thrown, perhaps from the use of |discontinue|. This
obviates the need for |dynamic-wind|.

OCaml does not have a \texttt{\footnotesize try/finally} construct available in
many programming languages. In OCaml, alternative standard libraries such as
Base~\cite{} and Core~\cite{} provide mechanisms analogous to Common Lisp
|unwind-protect|, which are in turn implemented using OCaml exceptions.
Leijen~\cite{Leijen18} explicitly extends effect handlers with |finally| clause
in Koka. Instead, Multicore OCaml |discontinue| primitive builds on extant
defensive coding practices against exceptions to clean up resources. This
design choice is crucial to ensure backwards compatibility of legacy systems
code with non-local control flow introduced by effect handlers.
Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
handlers and asynchronous exceptions. This is orthogonal to the contributions
of this paper which focusses on compiler and runtime system support for
implementing effect handlers. Ahman et al.~\cite{Ahman20} present a effect
handler core calculus with asynchrony which might help lay the formal
foundations for reconciling asynchrony in systems and effect handlers.

@@

We presented the motivation, design and implementation of effect handlers
specifically aimed at retrofitting them onto OCaml with tool compatibility,
negligible overheads for existing OCaml programs and good performance for those
that use effect handlers. Our evaluation shows that this is indeed the case.

>>>>

Our work is also timely. WebAssembly~\cite{Hass17} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\cite{WasmProposal}. Project Loom~\cite{loom} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{swift} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we observed
that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
cache.

@@

We consider well-behaved Multicore OCaml programs to resume their continuations
\emph{exactly once}. This can be enforced by installing a finaliser on every
captured continuation that |discontinue|s the continuation during finalisation
and ignores the result:
\begin{lstlisting}
Gc.finalise (fun k ->
  try ignore (discontinue k Unwind) | _ -> ()) k
\end{lstlisting}
Multicore OCaml does not enforce this behaviour in order to avoid installing an
effect handler every time a continuation is captured.

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is considered to be complex and not quite
what one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main
challenge is distinguishing returning exits from non-returning ones. Resource
clean up needs to occur only during non-returning exit.

In Multicore OCaml, we build on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code with non-local
control flow introduced by effect handlers.

Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
|finally| clauses in Koka for resource safety. Sivaramakrishnan et
al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
exceptions. This is orthogonal to the contributions of this paper. Our focus i
the compiler and runtime system support for implementing effect handlers.

@@

We presented the motivation, design and implementation of effect handlers
specifically aimed at retrofitting them onto OCaml with tool compatibility,
negligible overheads for existing OCaml programs and good performance for those
that use effect handlers. Our evaluation shows that this is indeed the case. We
plan to upstream the compiler and runtime support for effect handlers followed
by an effect system.
